---
title: An Introduction to R
minutes: 45
output:
  html_document:
    toc: true
    toc_depth: 1
---

```{r, echo=FALSE}
# The default setting for this lesson is to hide results. This is so
# the students will evaluate the code on their own computer.
knitr::opts_chunk$set(results = "hide", fig.path = "../img/r-lesson-")
```


# Learning Objectives

*   Familiarize participants with R syntax
*   Get comfortable working at the console
*   Understand the concepts of objects and assignment
*   Gain exposure to functions
*   Understand the concepts of vector and data types


# R Syntax: An Example Script

Below is a script to demonstrate some of the syntax of the R language.

```{r}
# Define a new function.
suggest <- function(n) {
    for (i in 1:n) {
        my_message <- "R is awesome!\n"
        cat(my_message)
    }
}

# Call that function.
suggest(n = 10)

# Create a new list object.
person <- list(first_name = "John", last_name = "Chambers")

# Print the person's first name.
cat(person$first_name, "\n")
```

To run this code:

1.  Open RStudio.
2.  Navigate to "File > New File > RScript".
3.  Copy and paste the code above into your untitled R Script.
4.  Press the "Source" button.

Let's take a quick walk through the code:

1.  Define a new function named `suggest` that takes a single argument `n`. `n` is the number of times this function will suggest R is awesome!

2.  The line `my_message <- "R is awesome!\n"` creates a new text variable called `my_message`. That line uses the assignment operator `<-` to assign a value to an object.

3.  Notice how the code is indented, this makes it easier for humans to read and understand the code structure.

4.  The line `suggest(n = 5)` calls the suggest function with an argument of `n = 5`. Although one can use `=` for assignment it is preferred to `<-` for assignment and `=` when passing arguments to a function.

5.  Notice that elements in a list can be accessed using the `$` operator. In this example, `person$first_name` equals "John". Fun fact, John Chambers created R when he was at Bell Labs.


# The Console

You can get output from R simply by typing math in the console:

```{r}
12 / 7
```

We can also comment what it is we're doing:

```{r}
# I am adding 3 and 5. R is fun!
3 + 5
```

What happens if we do that same command without the `#` sign in the front?

```{r, eval = FALSE}
I am adding 3 and 5. R is fun?
3 + 5
```

Now R is trying to run that sentence as a command, and it doesn't work.

What if we replace the hash `#` with a double quote `"`:

```{r, eval = FALSE}
" I am adding 3 and 5. R is fun?
3 + 5
```

Now we're stuck over in the console. The `+` sign means that it's still waiting for input, so we can't type in a new command. To get out of this type `Esc`. This will work whenever you're stuck with that `+` sign.

It's great that R is a glorified caluculator, but obviously we want to do more interesting things.

To do useful and interesting things, we need to assign _values_ to _objects_. To create objects, we need to give it a name followed by the assignment operator `<-` and the value we want to give it.


# Objects and Assignment

For instance, instead of adding `3 + 5`, we can assign those values to objects and then add them.

```{r}
# Assign 3 to a.
a <- 3
# Assign 5 to b.
b <- 5

# What is a now?
a
# What is b now?
b

# Add a and b.
a + b
```

`<-` is the assignment operator. It assigns values on the right to objects on the left. So, after executing `x <- 3`, the value of `x` is `3`. The arrow can be read as `3` goes into `x`. You can also use `=` for assignments but not in all contexts so it is good practice to use `<-` for assignments. `=` should only be used to specify the values of arguments in functions.

In RStudio, typing `Alt -` (push `Alt`, the key next to your space bar at the same time as the `-` key) will write `<-` in a single keystroke.

When assigning a value to an object, R does not print anything. You can force R to print the object's value by using parentheses:

```{r}
(s <- "hi")
```

or by typing the object's name:

```{r}
s <- "hi"
s
```

## Exercises

-   What happens if we change the value of `a` and then re-add `a` and `b`?

-   Does it work if you just change `a` in the script and then add `a` and `b`? Did you still get the same answer after they changed `a`? If so, why do you think that might be?

-   We can also assign `a + b` to a new variable, `c`. How would you do this?

## Notes on Objects

Objects can be given any name such as `x`, `current_temperature`, or `subject_id`. You want your object names to be explicit and not too long. They cannot start with a number (`2x` is not valid but `x2` is). R is case sensitive (`length_mb` is different from `length_MB`). There are some names that cannot be used because they represent the names of fundamental functions in R (`if`, `else`, `for`, ...). See [here][reserved] for a complete list of reserved keywords. In general, even if it's allowed, it's best to not use other function names (`c`, `T`, `mean`, `data`, `df`, `weights`, ...). In doubt check the help to see if the name is already in use. It's also best to avoid dots (`.`) within a variable name as in `my.dataset`. There are many functions in R with dots in their names for historical reasons, but because dots have a special meaning in R (for methods) and other programming languages, it's best to avoid them. It is also recommended to use nouns for variable names, and verbs for function names. It's important to be consistent in the styling of your code (where you put spaces, how you name variables, ...). In R, two popular style guides are [Hadley Wickham's][hadley] and [Google's][google].

[reserved]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html
[hadley]: http://adv-r.had.co.nz/Style.html
[google]: https://google-styleguide.googlecode.com/svn/trunk/Rguide.xml


# Functions and Function Arguments

The other key feature of R are functions. R has many built-in functions. Some examples of these are mathematical functions, like `sqrt` and `round`. You can also get functions from libraries (which we'll talk about in a bit), or even write your own.

Functions are "canned scripts" that automate something complicated or convenient or both. Many functions are predefined, or become available when using the function `library()` (more on that later). A function usually gets one or more inputs called arguments. Functions often (but not always) return a value. A typical example would be the function `sqrt`. The input (the argument) must be a number, and the return value (the output) is the square root of that number. Executing a function (running it) is called *calling* the function. An example of a function call is:

```{r}
a <- 2
sqrt(a)
```

Here, the value of `a` is given to the `sqrt` function, the `sqrt` function calculates the square root. This function is very simple, because it takes just one argument.

The return 'value' of a function need not be numerical (like that of `sqrt`), and it also does not need to be a single item: it can be a set of things, or even a data set. We'll see that when we read data files in to R.

Arguments can be anything, not only numbers or filenames, but also other objects. Exactly what each argument means differs per function, and must be looked up in the documentation (see below). If an argument alters the way the function operates, such as whether to ignore 'bad values', such an argument is sometimes called an *option*.

Most functions can take several arguments, but many have so-called *defaults*. If you don't specify such an argument when calling the function, the function itself will fall back on using the *default*. This is a standard value that the author of the function specified as being "good enough in standard cases". An example would be what symbol to use in a plot. However, if you want something specific, simply change the argument yourself with a value of your choice.

Let's try a function that can take multiple arguments `round`.

```{r}
round(3.14159)
```

We can see that we get `3`. That's because the default is to round to the nearest whole number. If we want more digits we can see how to do that by getting information about the `round` function. We can use `args(round)`:

```{r}
args(round)
```

or look at the help for this function using `?round` or `help(round)`:

```{r, eval=FALSE}
?round
help(round)
```

We see that if we want a different number of digits, we can type `digits = 2` or however many we want.

```{r}
round(3.14159, digits = 2)
```

If you provide the arguments in the exact same order as they are defined you don't have to name them:

```{r}
round(3.14159, 2)
```

However, it's usually not recommended practice because it's a lot of remembering to do, and if you share your code with others that includes less known functions it makes your code difficult to read. (It's however OK to not include the names of the arguments for basic functions like `mean`, `min`, ...).

Another advantage of naming arguments, is that the order doesn't matter. This is useful when there start to be more arguments. 

## Exercises

We're going to work with genome lengths.

- Create a variable `genome_length_mb` and assign it the value `4.6`

```{r}
(genome_length_mb <- 4.6)
genome_length_mb
```

Now that R has `genome_length_mb` in memory, we can do arithmetic with it. For instance, we may want to convert this to the weight of the genome in picograms (for some reason). 978Mb = 1picogram. Divide the genome length in Mb by 978.

```{r}
genome_length_mb / 978.0
```

It turns out an E. coli genome doesn't weigh very much.

We can also change the variable's value by assigning it a new one. Say we want to think about a human genome rather than E. coli. Change `genome_length_mb` to 3000 and figure out the weight of the human genome.

```{r}
genome_length_mb <- 3000.0
genome_length_mb / 978.0
```

This means that assigning a value to one variable does not change the values of other variables. For example, let's store the genome's weight in a variable.

```{r}
genome_weight_pg <- genome_length_mb / 978.0
```

and then change `genome_length_mb` to 100.

```{r}
genome_length_mb <- 100
```

What do you think is the current content of the object `genome_weight_pg`? 3.06 or 0.102?


# Vectors and Data Types

```{r, echo=FALSE}
# Vectors and data types
```

A vector is the most common and basic data structure in R, and is pretty much
the workhorse of R. It's basically just a list of values, mainly either numbers or
characters. They're special lists that you can do math with. You can assign this list of values to a variable, just like you
would for one item. For example we can create a vector of genome lengths:

```{r}
glengths <- c(4.6, 3000, 50000)
glengths
```

A vector can also contain characters:

```{r}
species <- c("ecoli", "human", "corn")
species
```

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r}
length(glengths)
length(species)
```

You can also do math with whole vectors. For instance if we wanted to multiply the genome lengths of all the genomes in the list, we can do

```{r}
5 * glengths
```

or we can add the data in the two vectors together

```{r}
new_lengths <- glengths + glengths
new_lengths
```

This is very useful if we have data in different vectors that we 
want to combine or work with. 

There are few ways to figure out what's going on in a vector. 

`class()` indicates the class (the type of element) of an object:

```{r}
class(glengths)
class(species)
```

The function `str()` provides an overview of the object and the elements it
contains. It is a really useful function when working with large and complex
objects:

```{r}
str(glengths)
str(species)
```

You can add elements to your vector simply by using the `c()` function:

```{r}
lengths <- c(glengths, 90) # adding at the end
lengths <- c(30, glengths) # adding at the beginning
lengths
```

What happens here is that we take the original vector `glengths`, and we are
adding another item first to the end of the other ones, and then another item at
the beginning. We can do this over and over again to build a vector or a
dataset. As we program, this may be useful to autoupdate results that we are
collecting or calculating.

We just saw 2 of the 6 **data types** that R uses: `"character"` and
`"numeric"`. The other 4 are:

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's an integer)
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
  `1+4i`) and that's all we're going to say about them
* `"raw"` that we won't discuss further

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`) and
factors (`factor`).
