---
title: Data Read/Write
slug: data-io
author: Data Carpentry contributors, Michael Levy, Bob Freeman, & Andrew Marder
minutes: 20
weight: 20
---

```{r, echo = FALSE}
source("setup.R")
```

There are a ton of ways to get data into and out of R. This lesson aims to give you the tools to navigate your many options. [RStudio's Data Import Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-import-cheatsheet.pdf) provides an excellent overview. Working in the social sciences you are likely to come across the following types of data:

```{r, results = "asis", echo = FALSE}
my_table <- data.frame(
    "Data Type" = c(
        "CSV",
        "SPSS, Stata, or SAS",
        "Excel",
        "Databases",
        "Web APIs",
        "Web Scraping"
    ),
    "Suggested Library" = c(
        "[readr](https://cran.r-project.org/web/packages/readr/)",
        "[haven](https://cran.r-project.org/web/packages/haven/)",
        "[readxl](https://cran.r-project.org/web/packages/readxl/)",
        "[DBI](https://cran.r-project.org/web/packages/DBI/)",
        "[httr](https://cran.r-project.org/web/packages/httr/)",
        "[rvest](https://cran.r-project.org/web/packages/rvest/)"
    )
)

knitr::kable(my_table, format = "html", col.names = c("Data Type", "Suggested Library"))
```

To keep this lesson manageable we will focus primarily on reading and writing .csv files, and we will show how working with Stata .dta files is quite similar.


## Paths to Files and Folders

One of the trickiest parts of working with R is telling it where to find files on disk. I found this [introduction to working with files and folders](http://www.dummies.com/programming/r/how-to-work-with-files-and-folders-in-r/) quite helpful. Here's a quick overview of that article.

Whenever you are working in R, your R session has a current working directory. This is the folder on your hard drive where R is working from. So, if you tell R to open a file named `my-data.csv` it will look in the current working directory to see if it can find that file. If that file doesn't exist in that directory you'll receive an error message.

```{r, error = TRUE}
read.csv("my-data.csv")
```

Often it is helpful to know where R is working from, to see the current working directory use the `getwd()` function:

```{r}
getwd()
```

As R compiles this lesson, it's working from the directory listed above. This is likely different from the output you see running the same function on your computer. To change the working directory use the `setwd()` function. Unfortunately paths on Windows are a little different compared to paths on Mac and Linux. RStudio makes navigating your file system a little easier. Use the "Files" pane to navigate into the folder where you want to be working, click the "More" button and select "Set As Working Directory". You'll see that RStudio calls the `setwd()` function with the appropriate path for your operating system.


## Read from CSV

There are many options for reading data from a CSV file. The two options I use most are:

1.  Base R's `read.csv()` function, and
2.  readr's `read_csv()` function.

I use `read.csv()` because I'm old and I remember the days when readr didn't exist. You're young so it makes sense to jump into using `readr::read_csv()`, it's better.

### Read a CSV 

The `continents` data.frame was useful for learning because it was so small, but it's time to graduate to something more interesting and realistic. Data come in many forms, and we need to be able to load them in R. For our own use and with others who use R, there are R-specific data structures we can use, like the .RDA file-type we just saw, but we need to be able to work with more general data types too. Comma-separated value (csv) tables are perhaps the most universal data structure. 

The gapminder dataset provides country-by-year level data on gdp, population, and longevity. I downloaded it and put it in the data directory of my project. You will do the same in a minute.

We can read csv's with the `read.csv` function. The argument to `read.csv` is the location of the file, relative to your working directory. Since I saved the gapminder data to the `data` directory of my project, I can load it with this. Note the use of tab-completion to find the file and get it exactly right without typos. 

```{r, eval=TRUE, results='hide'}
read.csv('data/gapminder-FiveYearData.csv')
```

Whoa! What just happened? R executed the function and printed the result, just like when you enter `log(1)`. How do you store an object to a variable?

```{r}
gapminder <- read.csv('data/gapminder-FiveYearData.csv')
```

This statement doesn't produce any output because assignment doesn't display
anything. But now, a data.frame called `gapminder` is in my Environment, and I can see that it is a data.frame with 1704 rows and 6 columns. If we want to check that our data has been loaded, we can print the
variable's value: `gapminder`

Alternatively, wrapping an assignment in parentheses will perform the assignment
and display it at the same time.

```{r, eval=TRUE, results='hide', purl = FALSE}
(gapminder <- read.csv('data/gapminder-FiveYearData.csv'))
```

Wow... that was a lot of output. At least it means the data loaded properly. Let's check the top (the first 6 lines) of this `data.frame` using the function `head()`:

```{r}
head(gapminder)
```

We've just done two very useful things.
1. We've read our data in to R, so now we can work with it in R
2. We've created a data frame (with the read.csv command) the 
standard way R works with data. 

A `data.frame` can be created by the functions `read.csv()` or `read.table()`, and can import 
local files or those from the web. These are usually text files in the shape of data 
tables that you might often import open in Microsoft Excel. If you wish to import Excel files,
you'll need to use an add-in (package) call XLConnect (see [this article](https://www.datacamp.com/community/tutorials/r-tutorial-read-excel-into-r#gs.njZB8ZI) for more information
on using Excel files). We'll briefly touch on packages later.

> Challenge -- read csv data
> 
> Suppose you get a .csv file from a colleague in Europe. Because they use "," (comma) as a decimal separator, they use ";" (semi-colons) to separate fields. How can you read it into R? 
> 
> Feel free to use the web and/or R's helpfiles.

By default, `data.frame` converts (= coerces) columns that contain characters
(i.e., text) into the `factor` data type. Depending on what you want to do with
the data, you may want to keep these columns as `character`. To do so,
`read.csv()` and `read.table()` have an argument called `stringsAsFactors` which
can be set to `FALSE`:

```{r eval=TRUE}
gapminder <- read.csv('data/gapminder-FiveYearData.csv', stringsAsFactors = FALSE)
```

Let's now check the __str__ucture of this `data.frame` in more details with the
function `str()`:

```{r}
# get the dataframe back to the original load
gapminder <- read.csv('data/gapminder-FiveYearData.csv')
str(gapminder)
```

## Inspect Data

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a `data.frame`. There are additional functions that can be
useful in helping to assess the structure and content of your data. Note: most of these 
functions are "generic", they can be used on other types of objects besides a `data.frame`.

### How big is my data?

Three commands will allow you to see the shape of your data. `dim()` will return a vector 
with the number of rows in the first element, and
the number of columns as the second element (the __dim__ensions of the object)

```{r}
dim(gapminder)
```

You can also assess these independently using both the `nrow()` and `ncol()` functions:

```{r}
nrow(gapminder)
ncol(gapminder)
```


### What's in my data?

As we've already seen the `head()` command, what might you guess is the complementary command
to look at the bottom of the data?

```{r}
tail(gapminder)
````


### What variables are in my data?

If your dataset has column headers, this information can be retrieved with the `colnames()` command.
More colloquially, you can use `names()` if your object is a `data.frame`.

```{r}
colnames(gapminder)
````
```{r}
names(gapminder)
````

Likewise, the `rownames()` command will retrieve the row headers, if present. If not, this
command will simply return the row number:

```{r}
# head this so the list isn't long
head(rownames(gapminder), n = 24L)
````


-----

You've already seen the structure command `str()`, to inspect the composition of the data object:

```{r}
str(gapminder)
````

Equally as valuable is the `summary()` command, which will give you summary statistics 
for each column:

```{r}
summary(gapminder)
````

As you can see, this information is more valuable for some columns than others.  **CHALLENGE!**

> Challenge
> 
> Based on the functions you've learned to asses data structure, can you answer the following questions?
> 
> * What is the class of the column `continent`?
> * The summary() function doesn't make sense for the whole gapminder object (variable). What command
> would you use to get a summary of the population?

As you can see, many of the columns in our data frame are of a special class called
`factor`. Before we learn more about the `data.frame` class, we are going to
talk about factors. They are very useful but not necessarily intuitive, and
therefore require some attention.


## Write to CSV


## Create Data in Code

You can create a `data.frame` manually with the function `data.frame()`.

```{r}
animals <- data.frame(
    name = c("Andrew", "Bob", "Carl", "Doug"),
    feel = c("Furry", "Furry", "Squishy", "Spiny"),
    type = c("Dog", "Cat", "Sea Cucumber", "Sea Urchin"),
    weight_lbs = c(45, 8, 1.1, 0.8)
)
```

Let's look at the `data.frame` we've created.

```{r}
animals
```

We have created a table of animals. There are four rows (animals) and four columns (characteristics of each animal). Andrew is a furry dog that weighs 45 pounds.

If you want to see the data type of each column or you're working with a large data frame and you don't want to print all of the data, the `str()` function is very useful for inspecting the structure of a data frame.

```{r}
str(animals)
```

## Stata .dta Files

The `data.frame()` function is not the only way to create `data.frame` objects. In this section we use the `load()` and `read.csv()` functions to create new data frames.

R comes with functions to `save()` objects to files and `load()` objects that have been saved to files. R uses its own binary file format. Files saved in this manner typically have a file extensions of `.RData`, `.rda`, or `.RDA`. Multiple objects can be saved to the same R data file, the `load()` function loads all the objects stored in the file. To see what objects have been loaded set the `verbose` argument to `TRUE`. Let's load a `data.frame` of continents:

```{r}
load('data/continents.RDA', verbose = TRUE)
```

This creates a new `data.frame` object called `continents`. Let's take a look at the data:

```{r}
continents
```

There are 6 rows and 4 columns of data in this table of continents. Notice how much easier it was to read a `data.frame` from a file than to create it manually.

Now let's create a `data.frame` from data contained in a CSV (Comma-Separated Values) file. This is one of most common ways to import data contained in a spreadsheet into R.

```{r}
gapminder <- read.csv("data/gapminder-FiveYearData.csv")
```

This data file is much larger than the data frames we have been working with. It has `r nrow(gapminder)` rows, so printing all the data would be a little messy. The `str()` function is useful for looking at the structure of a `data.frame`.

```{r}
str(gapminder)
```

By default, `data.frame()` converts (= coerces) columns that contain characters
(i.e., text) into the `factor` data type. Depending on what you want to do with
the data, you may want to keep these columns as `character`. To do so,
`read.csv()` and `read.table()` have an argument called `stringsAsFactors` which
can be set to `FALSE`.
