---
title: Data Read/Write
slug: data-io
author: Data Carpentry contributors, Michael Levy, Bob Freeman, & Andrew Marder
minutes: 20
weight: 20
---

```{r, echo = FALSE}
source("setup.R")
```

There are many ways to get data into and out of R. This lesson aims to give you the tools to navigate your many options. [RStudio's Data Import Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-import-cheatsheet.pdf) provides an excellent overview. Working in the social sciences you are likely to come across the following types of data:

```{r, results = "asis", echo = FALSE}
my_table <- data.frame(
    "Data Type" = c(
        "CSV",
        "SPSS, Stata, or SAS",
        "Excel",
        "Databases",
        "Web APIs",
        "Web Scraping"
    ),
    "Suggested Library" = c(
        "[readr](https://cran.r-project.org/web/packages/readr/)",
        "[haven](https://cran.r-project.org/web/packages/haven/)",
        "[readxl](https://cran.r-project.org/web/packages/readxl/)",
        "[DBI](https://cran.r-project.org/web/packages/DBI/)",
        "[httr](https://cran.r-project.org/web/packages/httr/)",
        "[rvest](https://cran.r-project.org/web/packages/rvest/)"
    )
)

knitr::kable(my_table, format = "html", col.names = c("Data Type", "Suggested Library"))
```

To keep this lesson manageable we will focus primarily on reading and writing .csv files, and we will show how working with Stata .dta files is quite similar.


## Paths to Files and Folders

One of the trickiest parts of working with R is telling it where to find files on disk. I found this [introduction to working with files and folders](http://www.dummies.com/programming/r/how-to-work-with-files-and-folders-in-r/) quite helpful. Here's a quick overview of that article.

Whenever you are working in R, your R session has a current working directory. This is the folder where R is working from. So, if you tell R to open a file named `my-data.csv` it will look in the current working directory to see if it can find that file. If that file does not exist in that directory you'll receive an error message.

```{r, error = TRUE}
library(readr)

read_csv("my-data.csv")
```

Often it is helpful to know where R is working from, to see the current working directory use the `getwd()` function:

```{r}
getwd()
```

As R compiles this lesson, it's working from the directory listed above. This is likely different from the output you see running the same function on your computer. To change the working directory use the `setwd()` function. Unfortunately paths on Windows are a little different compared to paths on Mac and Linux. RStudio makes navigating your file system a little easier. Use the "Files" pane to navigate into the folder where you want to be working, click the "More" button and select "Set As Working Directory". You'll see that RStudio calls the `setwd()` function with the appropriate path for your operating system.

### Tab-Completion of Paths

Note RStudio provides tab-completion of paths to find files and avoid typos.


## Read from CSV

> A comma-separated values (CSV) file stores tabular data (numbers and text) in plain text. Each line of the file is a data record. Each record consists of one or more fields, separated by commas. The use of the comma as a field separator is the source of the name for this file format.

Download the data for this lesson from [here](/R_Intro-gapminder/data/gapminder-FiveYearData.csv). The gapminder dataset provides country-by-year level data on GDP, population, and life expectancy.

Use RStudio's "Files" pane to navigate to where you downloaded the data.

Read the CSV with the `read_csv()` function from the readr package. The argument to `read_csv()` is the location of the file, relative to your working directory. Run the following code to read the data into R:

```{r}
library(readr)

gapminder <- read_csv("gapminder-FiveYearData.csv")
```

If readr is not installed on your computer, use `install.packages("readr")` to install it.

We have created a new object in the global environment called `gapminder`. Looking at RStudio's "Environment" pane we see this object has type `r class(gapminder)[1]`, it has `r ncol(gapminder)` columns and `r nrow(gapminder)` rows. If we want to check that our data has been loaded, we can print the variable's value:

```{r}
gapminder
```

Congratulations! You have successfully read data into R.


## Inspect Data

Now that you have data in R, let's run through a few functions that are helpful for inspecting your data.

Use the `head()` function to see the first few rows of a data frame.

```{r}
head(gapminder)
```

Use the `tail()` function to see the last few rows:

```{r}
tail(gapminder)
```

Use the `str()` function to inspect the __str__ucture of an object:

```{r}
str(gapminder, give.attr = FALSE)
```

Note: If you use the `read.csv()` function from base R, the default behavior is to convert (= coerce) character columns to be stored as factors (integer values with character labels). Using the `read_csv()` function from readr avoids this issue (the country and continent columns are stored as character vectors).

```{r}
str(read.csv("gapminder-FiveYearData.csv"))
```

### How big is my data?

Three commands will allow you to see the shape of your data. `dim()` will return a vector 
with the number of rows in the first element, and
the number of columns as the second element (the __dim__ensions of the object)

```{r}
dim(gapminder)
```

You can also assess these independently using the `nrow()` and `ncol()` functions:

```{r}
nrow(gapminder)
ncol(gapminder)
```

### What variables are in my data?

If your dataset has column headers, this information can be retrieved with the `colnames()` command.
More colloquially, you can use `names()` if your object is a `data.frame`.

```{r}
colnames(gapminder)
```

```{r}
names(gapminder)
```

Likewise, the `rownames()` command will retrieve the row headers, if present. If not, this command will simply return the row number:

```{r}
# look at the head of all the rownames
head(rownames(gapminder))
````

### How are the variables distributed?

The `summary()` function is a very convenient way to get summary statistics for each column:

```{r}
summary(gapminder)
````

As you can see, this information is more valuable for some columns than others. We can use the `table()` function to tabulate categorical variables.

```{r}
table(gapminder$continent)
```


## Write to CSV


## Create Data in Code

You can create a `data.frame` manually with the function `data.frame()`.

```{r}
animals <- data.frame(
    name = c("Andrew", "Bob", "Carl", "Doug"),
    feel = c("Furry", "Furry", "Squishy", "Spiny"),
    type = c("Dog", "Cat", "Sea Cucumber", "Sea Urchin"),
    weight_lbs = c(45, 8, 1.1, 0.8)
)
```

Let's look at the `data.frame` we've created.

```{r}
animals
```

We have created a table of animals. There are four rows (animals) and four columns (characteristics of each animal). Andrew is a furry dog that weighs 45 pounds.

If you want to see the data type of each column or you're working with a large data frame and you don't want to print all of the data, the `str()` function is very useful for inspecting the structure of a data frame.

```{r}
str(animals)
```

## Stata .dta Files

The `data.frame()` function is not the only way to create `data.frame` objects. In this section we use the `load()` and `read.csv()` functions to create new data frames.

R comes with functions to `save()` objects to files and `load()` objects that have been saved to files. R uses its own binary file format. Files saved in this manner typically have a file extensions of `.RData`, `.rda`, or `.RDA`. Multiple objects can be saved to the same R data file, the `load()` function loads all the objects stored in the file. To see what objects have been loaded set the `verbose` argument to `TRUE`. Let's load a `data.frame` of continents:

```{r}
load('continents.RDA', verbose = TRUE)
```

This creates a new `data.frame` object called `continents`. Let's take a look at the data:

```{r}
continents
```

There are 6 rows and 4 columns of data in this table of continents. Notice how much easier it was to read a `data.frame` from a file than to create it manually.

Now let's create a `data.frame` from data contained in a CSV (Comma-Separated Values) file. This is one of most common ways to import data contained in a spreadsheet into R.

```{r}
gapminder <- read.csv("gapminder-FiveYearData.csv")
```

This data file is much larger than the data frames we have been working with. It has `r nrow(gapminder)` rows, so printing all the data would be a little messy. The `str()` function is useful for looking at the structure of a `data.frame`.

```{r}
str(gapminder)
```

By default, `data.frame()` converts (= coerces) columns that contain characters
(i.e., text) into the `factor` data type. Depending on what you want to do with
the data, you may want to keep these columns as `character`. To do so,
`read.csv()` and `read.table()` have an argument called `stringsAsFactors` which
can be set to `FALSE`.
